---
title: "Session 3 - Visualization"
author: "Jongbin Jung"
date: "January 9-10, 2016"
output: 
  beamer_presentation: 
    fig_height: 3
    fig_width: 3
    fonttheme: professionalfonts
    highlight: zenburn
    theme: boxes
---

## Dependencies
- Latest version ($\ge$ `3.1.2`) of `R`  
  (*free* from <https://www.r-project.org/>)
- Latest version of `Rstudio` 
  (also *free* from <https://www.rstudio.com/>)
- A bunch of *free* packages

```{r, eval=FALSE}
install.packages('ggplot2')
```
```{r, echo=FALSE, results='hide', message=FALSE}
options(width=50)
library('ggplot2')
```

## Visualization: Introduction
- There is more than one framework for thinking about data visualization, e.g.,
    1. Mapping of vectors to 2D/3D surfaces 
    2. Function of **inputs** that are given as varaibles of a data set, 
    **geometries** and **aesthetics** that describe visual markings, and a 
    **coordinate** system that defines the location of each marking
    
- The first approach is widely used in scientific visualization (e.g., `MATLAB`, 
classical plotting function in `R`),  but doesn't scale well with data
- It helps your sanity to always think of data frames as a table where
```
Each column represents a variable/feature
Each row represents an observation/instance
```

## Reference
- A great "cheat sheet" for data visualization with `ggplot2` is available
for free at 
https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf