---
title: "Session 4 - Statistical Modeling"
author: "Jongbin Jung"
date: "January 10, 2016"
output: 
  beamer_presentation: 
    fig_height: 4
    fig_width: 4
    fonttheme: professionalfonts
    highlight: zenburn
    theme: boxes
---

## Dependencies
- Latest version ($\ge$ `3.1.2`) of `R`  
  (*free* from <https://www.r-project.org/>)
- Latest version of `Rstudio` 
  (also *free* from <https://www.rstudio.com/>)
- Packages
```{r, eval=FALSE}
lm, glm # loaded with R by default
install.packages('glmnet')
install.packages('randomForest')
install.packages('ggplot2')  # just for dataset
install.packages('dplyr')
install.packages('caret')  # optional
# for evaluating classification models
install.packages('ROCR')  
```
- We'll use the `diamonds` dataset included in `ggplot2` for demonstration
```{r, echo=FALSE, results='hide', message=FALSE}
options(width=50, digits=3)
library('glmnet')
library('randomForest')
library('ggplot2')
library('dplyr')
library('caret')
```

## Statistical Modeling
- In this session, I'll share some statistical modeling techniques I've picked 
up using `R`
- But I *won't*
    - Go into the theory/details of each (or any) method/model; I assume you 
    know all that already
    - Introduce the whole `caret` package, which is a **great** package, but I
    won't focus on it because:
        1. it's pretty well documented [here](http://topepo.github.io/caret/index.html)
        (http://topepo.github.io/caret/index.html), and
        2. I (very personally) don't think 'teaching people how to  use the 
        `caret`  package' is  equivalent to 'teaching people statistical 
        modeling in `R`' 
- At the very least, you'll end up with a set of (hopefully useful) `R` snippets

## Basic Framework
- Just to refresh, a basic framework of the statistical modelling process looks 
something like,
    1. Split the data: Train/(Validation/Test) or CV
    2. Format the data appropriately (pre-processing)
    3. Train a model on the training data
    4. Evaluate the performance on validation/test data
- Would you agree with the order of steps 1 and 2?
- Some considerations in each step, let's discuss a few
        
# Data Splitting

## Random Sampling
- Most simple, but often sufficient way of splitting data is to generate random 
samples
- Think of generating a sample of row numbers, and then using the row numbers to 
actually create each dataset, e.g., for 50/50 split
```{r, eval=FALSE}
ind <- 1:nrow(diamonds)
train_p <- .5  # proportion of training data
train_ind <- sample(ind, train_p * nrow(diamonds))
diamond_train <- diamonds[train_ind, ]
diamond_test <- diamonds[-train_ind, ]
```

## Stratified Sampling
- Split data while maintaining proportion of certain subgroups
- Use `group_by()` and `sample_frac()` in `dplyr` to select a subset of the data
that satisfies the criteria
- Use `setdiff()` in `dplyr` to creat the complement subset

## Stratified Sampling: Example
```{r}
diamond_train <- diamonds %>% group_by(cut) %>%
    sample_frac(.5) %>% ungroup()
diamond_test <- setdiff(diamonds, diamond_train)

# check the proportions
cbind(test=summary(diamond_train$cut)/nrow(diamonds),
      train=summary(diamond_test$cut)/nrow(diamonds))
```

## More than One Split
- Often, you'll want more than one split, e.g., train/validate/test, 
cross validation
- One obvious way is to use the previous method recursively
- Let's try this as an **Exercise**!

1. From the `diamonds` data, create a 50:30:20 split of train:validate:test 
data. Name the data frames `dia_train`, `dia_valid`, and `dia_test`, 
respectively.

(solution script is on the next slide)

## More than One Split: Exercise Solution
```{r}
ind <- 1:nrow(diamonds)
train_p <- .5  # proportion of training data
valid_p <- .3  # proportion of validation data
train_ind <- sample(ind, train_p * nrow(diamonds))
dia_train <- diamonds[train_ind, ]
dia_tmp <- diamonds[-train_ind, ]
ind <- 1:nrow(dia_tmp)
valid_ind <- sample(ind, valid_p * nrow(diamonds))
dia_valid <- dia_tmp[valid_ind, ]
dia_test <- dia_tmp[-valid_ind, ]
```
- We'll use these three datasets in the following exercises

## More than One Split (cont'd)
- As you can imagine, this starts getting messy for more than two splits
- A good alternative is `createFolds()` from `caret` (I know I said I wouldn't 
cover `caret`, but this is one exception ... )
- Also, for more than 3 splits, you might want to manage each split with labels, 
rather than creating multiple data frames
```{r, eval=FALSE}
nsplits <- 10  # the number of splits you want
split_ind <- createFolds(diamonds$carat, k=nsplits)

diamonds_split <- diamonds
for (x in 1:nsplits) {
    ind <- split_ind[[x]]  # indexing a list
    diamonds_split[ind, 'split_id'] <- x
}
```

# Pre-processing

## `scale()`
- Use `scale()` to center/scale variables (columns) of your dataset
- `scale()` only works on numerical columns
- It's up to you to give `scale()` just the variables you want to manipulate
- The general idea is
    1. Extract the variable(s) you want to center/scale
    2. Use `scale()` to manipulate those variables
    3. Create a copy of your original data with the desired variables 
    manipulated
- Remember to center/scale **all** partitions of your data, **but be aware of
where the centering/scaling parameters come from!**

## `scale()`: Example
```{r}
# Create a copy of the data 
train_std <- diamond_train
test_std <- diamond_test

# extract numerical columns and their names
train_num <- train_std[, sapply(train_std, is.numeric)]
test_num <- test_std[, sapply(test_std, is.numeric)]
numcol_names <- names(train_num)

# apply scale() to train data and save parameters
train_num <- scale(train_num)
param_center <- attr(train_num, 'scaled:center')
param_scale <- attr(train_num, 'scaled:scale')
```

## `scale()`: Example (cont'd)
```{r}
# apply to numeric test data columns
test_num <- scale(test_num, center=param_center,
                  scale = param_scale)

# replace numeric columns with scaled ones
train_std[, numcol_names] <- train_num
test_std[, numcol_names] <- test_num
```
- Some notes:
    - Be careful about how you choose 'numeric' variables: binary variables?
    - There are other ways to do this, but this seems to be the best I've found 
    so far 

## `model.matrix()`
- While many models work just fine with data frames, some models require that 
you provide data in the form of a purely numeric matrix (aka model matrix)
- This means converting `factor` variables into multiple binary variables 
(variables that only have 0 or 1 as values)
- The `model.matrix()` function in `R` does a good job of generating model 
matrices catered to the formula of your model
- The `R` representation of a model formula such as

$$y_{\mathtt{carat}} = \beta_0 + \beta_{\mathtt{cut}} x_{\mathtt{cut}} +
    + \beta_{\mathtt{depth}} x_{\mathtt{depth}}$$

would be

```{r, eval=FALSE}
carat ~ cut + depth
```

## `model.matrix()`: Example
- To construct a model matrix for the formula
$$y_{\mathtt{carat}} = \beta_0 + \beta_{\mathtt{cut}} x_{\mathtt{cut}} +
     \beta_{\mathtt{depth}} x_{\mathtt{depth}}$$
    
```{r}
train_mm <- model.matrix(carat ~ cut + depth, 
                         train_std)
head(train_mm, 2)
levels(train_std$cut)
```

## `model.matrix()`: Example (cont'd)
- Note that
    - Orthogonal polynomial coding is used for ordinal variable `cut`, where
    `.L`, `.Q`, `.C`, and `^4` stand for Linear, Quadratic, Cubic, and 4th power
    - `model.matrix()` drops one level as the 'base case', c.f., `cut` has five
    levels but only four orders in the `model.matrix`

- Some shortcuts in formula
    - "`.`" is used to include all variables (except the target, i.e., variable 
    to the left of `~`)
    - "`:`" is used to indicate interaction terms
    - "`-`" (as opposed to `+`) can be used to exclude certain variables
    
## Exercise
1. With the datasets `dia_train` and  `dia_test`, creat an 
additional variable `expensive`, which is a binary variable with  value `yes` 
if `price` is greater than the median of `price` from `dia_train`,  and `no` 
otherwise.
2. Standardize (scale and center) all numeric columns of the `dia_train` and 
`dia_test` datasets and call them `train_std` and `test_std`, respectively.
3. Generate model matrices that uses all variables except `expensive` to 
predict `price` for both datasets. Use variable names `train_mm` and `test_mm`.
Note we can use these datasets to train/test a model to predict `expensive` 
as well!

## Solution 1
```{r}
medprice <- median(dia_train$price)
dia_train <- dia_train %>%
    mutate(expensive = 
               ifelse(price > medprice, 'yes', 'no'))
dia_test <- dia_test %>%
    mutate(expensive = 
               ifelse(price > medprice, 'yes', 'no'))
```

## Solution 2
```{r}
train_std <- dia_train
test_std <- dia_test

train_num <- train_std[, sapply(train_std, is.numeric)]
test_num <- test_std[, sapply(test_std, is.numeric)]
numcol_names <- names(train_num)

train_num <- scale(train_num)
param_center <- attr(train_num, 'scaled:center')
param_scale <- attr(train_num, 'scaled:scale')

test_num <- scale(test_num, center=param_center,
                  scale = param_scale)

train_std[, numcol_names] <- train_num
test_std[, numcol_names] <- test_num
```

## Solution 3
```{r}
train_mm <- 
    model.matrix(price ~ . - expensive, train_std)
test_mm <- 
    model.matrix(price ~ . - expensive, test_std)
```

# Training models

## (OLS) Linear Regression
- Linear regression models can be fitted in `R` using `lm`, with the syntax
```{r, eval=FALSE}
my_model <- lm(formula, data)
```
- The `data` should be a data frame, and the formula should refer to the column 
names of `data` as variables

- Explore the model with generic functions `summary()` and `coef()`, 
once trained
```{r, eval=FALSE}
summary(my_model)
coef(my_model)
```
- Values (e.g., 'risiduals') of the model can be indexed with the `$` operator

## (OLS) Linear Regression: Example
- Let's build a linear regression model of `price` against `carat` and `cut`, 
using the `train_std` we created earlier
```{r}
fm <- lm(price ~ carat + cut, train_std)
```
- Explore the model with
```{r, eval=FALSE}
coef(fm)
summary(fm)
```
- Note how `cut` is automatically transformed to an appropriate form

## Logistic Regression with `glm`
- We can use `glm` for generalized linear models, just like we use `lm` for 
OLS models
```{r, eval=FALSE}
my_model <- glm(formula, data, family)
```
- `family` is a description of the link function to be used
- Recall, a logistic regression is a generalized linear model that uses a logit 
link function
- The logit link function is described in a `binomial` family object in `R` 
(see `?family` for other link functions)
- So, to fit a logistic regression model, we write
```{r, eval=FALSE}
my_model <- glm(formula, data, family='binomial')
```
- Keep in mind, the target variable in the formula must numeric between 0 and 1

## Logistic Regression with `glm`: Example
- With our `test_std` dataset, let's fit a logistic regression model of 
`expensive` against `carat` and `cut`
- Since `expensive` is not numeric, we must convert it first
```{r, eval=FALSE}
train_std$exp_numeric <- 
    ifelse(train_std$expensive == 'yes', 1, 0)
fm <- glm(exp_numeric ~ carat + cut, 
          train_std, family='binomial')
```
- Explore the model with
```{r, eval=FALSE}
coef(fm)
summary(fm)
```

## Regularized Linear Models with `glmnet`
- There are many packages that deal with regularized linear models
- I (personally) find `glmnet` to be most useful and consistent
- The objective function for `glmnet` is

$$ \mathtt{glm}~\mathrm{objective} - \lambda\times\mathtt{penalty} $$

- Where the `penalty` is defined as

$$ (1-\alpha)/2\Vert \beta \Vert_2^2 + \alpha \Vert \beta \Vert_1 $$

- Note that $\alpha=1$ is the L1 (lasso) penalty, and $\alpha=0$ is the L2 
(ridge) penalty

## Regularized Linear Models with `glmnet` (cont'd)
- The `glmnet` syntax is
```{r, eval=FALSE}
my_model <- glmnet(x, y, family, alpha, nlambda)
```

- **`x`** is the `model matrix`
- **`y`** is the target variable vector
- **`family`** determines the link function (Gaussian by default)
- **`alpha`** determines lasso (1), ridge (0), or a mix of the two
- **`nlambda`** controls the number of $\lambda$ values to try
- Note that `glmnet` automatically computes its own sequence of $\lambda$ values 
based on `nlambda`, and using a specific value of $\lambda$ is discouraged (see
`?glmnet` for details)

## Regularized Linear Models with `glmnet`: Choosing $\lambda$
- The penalty weight $\lambda$ is a free parameter in regularized linear models,
and must be determined somehow
- A quick and dirty, but pretty reasonable way to choose $\lambda$ with `glmnet`
is to use `cv.glmnet()`
- `cv.glmnet()` does k-fold cross-validation for `glmnet`, and returns 
values/models for $\lambda$
- **In addition to `glmnet` arguments**, in `cv.glmnet()` we can set the number of 
folds $k$ to use in k-fold cross validation by supplying the argument `nfolds`. 
Default is 10.
- The `plot()` function for `cv.glmnet()` provides some insight on the value of 
$\lambda$ and cross-validated model performance

## Regularized Linear Models: Example
- For L1-regularized model of `price` against `carat` and `cut`
```{r}
y <- train_std$price
x <- model.matrix(price ~ carat + cut, train_std)
price_l1 <- cv.glmnet(x, y, alpha=1)
```

- For L2-regularized model of `expensive` against `carat` and `cut`
```{r}
y <- ifelse(train_std$expensive == 'yes', 1, 0)
x <- model.matrix(expensive ~ carat + cut, train_std)
expensive_l2 <- 
    cv.glmnet(x, y, alpha=0, family='binomial')
```

## Regularized Linear Models: Example (plots)
```{r, fig.height=3.3}
plot(price_l1)
```

## Regularized Linear Models: Example (plots)
```{r, fig.height=3.3}
plot(expensive_l2)
```

## Random Forests
- The `randomForest` package in `R` implements Breiman's random forest 
algorithm, based on Breiman and Cutler's original Fortran code
- It can be used for classification, regression, or unsepervised learning of 
proximity between data points
- The basic syntax is
```{r, eval=FALSE}
randomForest(x, y, ntree)
```

- **`x`**: the model matrix
- **`y`**: target variable vector. *Classification* if `y` is a factor, 
*regression* if `y` is otherwise, and *unsupervised* learning if `y` is omitted.
- **ntree**: number of trees to include in the forest
- While `randomForest` will also work for the `formula`/data frame syntax, a 
model matrix is considered more efficient

## Random Forests: Examples
- To fit random forest of `price` against `carat` and `cut`
```{r, eval=FALSE}
y <- train_std$price
x <- model.matrix(price ~ carat + cut, train_std)
rf_price <- randomForest(x, y, ntree=100)
```

- To fit a random forest `expensive` against `carat` and `cut`
```{r, eval=FALSE}
y <- factor(train_std$expensive)
x <- model.matrix(expensive ~ carat + cut, train_std)
rf_expensive <- randomForest(x, y, ntree=100)
```

- `randomForest` is quite memory intensive
- `randomForest` classification aggregates *votes* of individual trees (as 
opposed to probability assessments), hence the floating-point precision is 
strictly determined by the order of `ntree`, i.e., 1/`ntree`

## Exercises
- Fit the following models, with `train_std` and `train_mm` where appropriate, 
from previous exercises
    - `lm_price`: OLS model of `price` against everything 
    - `logit_exp`: Logistic regression of `expensive` against everything 
    - `l1_price`: L1 regularized logistic regression of `price` against 
    everything
    - `rf_exp`: Random forest classification model with 50 trees that uses 
    `cut`, `color`,  and `clarity` to  predict `expensive`
- When I say "everything", I mean "everything but `expensive`" for `price`, and 
vice versa

## Solution
```{r, warning=FALSE}
lm_price <- lm(price ~ . - expensive, train_std)

train_std$expensive_num <- 
    ifelse(train_std$expensive == 'yes', 1, 0)
logit_exp <- 
    glm(expensive_num ~ . - price - expensive,  
        train_std, family='binomial')

l1_price <- 
    cv.glmnet(x=train_mm, y=train_std$price, alpha=1,
              standardize=FALSE, intercept=FALSE)

y <- factor(train_std$expensive)
x <- model.matrix(expensive ~ cut + color + clarity, train_std)
rf_exp <- randomForest(x, y, ntree=50)
```

# Prediction/Evaluation

## The Generic `predict()` 
- Any decent modeling package in `R` will provide its own version of the generic 
`predict()` function
- The generic syntax is
```{r, eval=FALSE}
predict(model_object, newdata, type)
```
- The `type` argument is often optional, and specifies the 'scale' of the 
predictions, e.g., probabilities v. log-odds for logistic regression models
- It's best (but not always necessary) for the `newdata` to have the same 
variables (columns) as the data that was used to train the model

## The Generic `predict()`: Examples
- Using the models from the previous exercise, we can generate predictions for 
the `test_std` data
```{r}
lm_price_pred <- predict(lm_price, test_std)

test_std$expensive_num <- 
    ifelse(test_std$expensive == 'yes', 1, 0)

logit_exp_pred <- predict(logit_exp, test_std, 
                          type='response')

l1_price_pred <- predict(l1_price, test_mm, 
                         s='lambda.min')

rf_exp_pred <- predict(rf_exp, test_mm, 
                       type='prob')
```

## Example Notes
- The `s` argument for `predict.cv.glmnet` specifies the criteria for selecting 
a $\lambda$ value from those that were fitted (see `?predict.cv.glmnet` for details)
- Beware of the column ordering for `rf_exp_pred` !
```{r}
head(rf_exp_pred, n=2)
```

## Manual Predictions
- For linear models, generating predictions from new data should be as simple as
matrix multiplication (with the help of `coef()`), e.g.,
```{r}
manual_l1_price <-
    cbind(1, test_mm) %*%
    coef(l1_price, s='lambda.min') 
all(manual_l1_price == l1_price_pred)
```
- What might *not* be simple is creating the right matrix
- It's probably safe to stick with `predict()`, but make **sure** you read the 
documentation for new packages!!!!

## Evaluation
- Once the predictions on the test set are generated, you should be able to 
compute any performance measure you're interested in
```{r}
# RMSE for lm_price
sqrt(mean((test_std$price - lm_price_pred)^2))
# RMSE for l1
sqrt(mean((test_std$price - l1_price_pred)^2))
```

## Evaluation (cont'd)
```{r}
# Accuracy for logit_exp (for threshold 0.5)
logit_exp_vote <- ifelse(logit_exp_pred >= .5, 1, 0)
mean(logit_exp_vote == test_std$expensive_num)
# Accuracy for rf_exp (for threshold 0.5)
rf_exp_vote <- ifelse(rf_exp_pred[,2] >= .5, 1, 0)
mean(rf_exp_vote == test_std$expensive_num)
```

## Evaluating Classifications
- The `ROCR` package can come in handy when evaluating classification models
```{r, message=FALSE, warning=FALSE}
library(ROCR)
pred <- prediction(rf_exp_pred[, 2], 
                   test_std$expensive_num)
auc <- performance(pred, 'auc')
unlist(slot(auc, 'y.values'))
```

## Evaluating Classifications (cont'd)
```{r, fig.height=3, fig.width=3}
perf <- performance(pred, 'tpr', 'fpr')
plot(perf)
```

- See website https://rocr.bioinf.mpi-sb.mpg.de/ for details and more examples

# Bootstrap

## The Sledgehammer
- The actual implementation of generating bootstrap sampling distributions is 
super simple
- What might not be as simple is
    1. Truly understanding the theory
    2. Finding the right way of constructing measure of interest, e.g., 
    confidence interval
    3. Implementing in a concise, reusable, and efficient way
- There are probably entire courses covering the first two
- People tend to gloss over the third
- I'll **only** focus on the third
- But keep in mind, knowing how to do it is very different from doing it right

## Framework
1. Write a dummy function that 
    - takes a single (dummy) argument
    - creates a single bootstrap sample from the original sample
    - returns the statistic of interest of the boostrap sample (as a scalar)
2. Use `boot <- sapply(1:B, dummy_function)` to create a bootstrap estimation 
of the  sampling distribution
3. Compute the measure of interest (e.g., confidence interval) using the 
original sample and the `boot` vector created in (2)

- To save computation time, let's use a subsample of `diamonds` for the 
following examples
```{r}
diamonds_small <- sample_n(diamonds, 5000)
```


## Simple Example
- Construct a bootstrap estimate of the sampling distribution for the mean of 
`price` in the `diamonds` data, with $B=1,000$
```{r, cache=TRUE}
# 1. A dummy function
get_mean <- function(x) {
    mean(sample(diamonds_small$price, 
                nrow(diamonds_small), 
                replace=TRUE))
}
# 2. Creating the boot vector
boot <- sapply(1:1000, get_mean)
```

## Histogram of `boot` and Original Point Estimate
```{r, echo=FALSE, cache=TRUE, fig.width=7}
qplot(boot, bins=20) + 
    geom_vline(aes(xintercept=mean(diamonds_small$price)), color='red', linetype=2)
```

## Less Simple Example
- Construct a bootstrap estimate of the sampling distribution for the 
coefficient of `carat` for a OLS model fitting `price` to everything

```{r, cache=TRUE}
# 1. A dummy function
get_coef <- function(x) {
    df <- sample_frac(diamonds_small, size=1,
                      replace=TRUE)
    fm <- lm(price ~ ., df)
    as.numeric(coef(fm)['carat'])
}
# 2. Creating the boot vector
boot <- sapply(1:1000, get_coef)
```

## Histogram of `boot` and Original Point Estimate
```{r, echo=FALSE, cache=TRUE, fig.width=7}
fm <- lm(price ~ ., diamonds_small)
point.est <- as.numeric(coef(fm)['carat'])
qplot(boot, bins=20) + 
    geom_vline(aes(xintercept=point.est), color='red', linetype=2)
```