---
title: "Session 4 - Statistical Modeling"
author: "Jongbin Jung"
date: "January 10, 2016"
output: 
  beamer_presentation: 
    fig_height: 2
    fig_width: 2
    fonttheme: professionalfonts
    highlight: zenburn
    theme: boxes
---

## Dependencies
- Latest version ($\ge$ `3.1.2`) of `R`  
  (*free* from <https://www.r-project.org/>)
- Latest version of `Rstudio` 
  (also *free* from <https://www.rstudio.com/>)
- Packages
```{r, eval=FALSE}
install.packages('glmnet')
install.packages('ROCR')
```
```{r, echo=FALSE, results='hide', message=FALSE}
options(width=50)
library('glmnet')
```

## Statistical Modeling
- In this session, I'll demonstrate some common statistical modeling 
techniques used in `R`
- But I *won't*
    - Go into the theory/details of each (or any) model/method; I assume you 
    know all that already
    - Introduce the `carat` package, which is a **great** package, but I'm not 
    dealing with it because:
        1. it's pretty well documented [here](http://topepo.github.io/caret/index.html)
        (http://topepo.github.io/caret/index.html), and
        2. I (very personally) don't think 'teaching people how to  use the 
        `carat`  package' is  equivalent to 'teaching people statistical 
        modeling in `R`' 
- I'll focus on regularized linear models with the `glmnet` package, but many of 
the general concepts work for most models/packages
- At the very least, you'll end up with a set of (hopefully useful) `R` snippets
        
# Data Splitting

# Pre-processing

## `model.matrix()`

## `scale()`

# Training models

# Prediction/Evaluation

# Bootstrap